---
title: "Project Part 1: Unemployment Patterns" 
fontsize: 12pt
geometry: margin=1in
urlcolor: black
output: pdf_document
header-includes:
- \usepackage{setspace}
- \onehalfspacing
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(kableExtra.latex.load_packages = FALSE)   # <- stop kableExtra from auto-loading (incl. tabu)
```



```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Loading and Renaming data sets
library(tidyverse)
library(kableExtra)

college<-read.csv("college.1.csv")
non.college <- read.csv("highschool.csv")

colnames(college) <- c("Month", "UR_College")
colnames(non.college) <- c("Month", "UR_NonCollege")

# Changing the months column to display month names
college$Month <- month.name[as.numeric(substr(college$Month, 6,7))]
non.college$Month <- month.name[as.numeric(substr(non.college$Month, 6,7))]

# Merging data
unemployment.data <- merge(college, non.college, by="Month")
unemployment.data$Month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug",
                             "Sep", "Oct", "Nov", "Dec")
```

## Introduction

  Unemployment is one of the most widely discussed indicators of a country’s economic health and labor market efficiency. It reflects a country’s economic performance, stability, and the well-being of households, communities, and future workforce participants. In recent years, the U.S. labor market has undergone substantial changes as it continues to recover from the pandemic’s disruption and adjust to new patterns in remote work, automation, and sectoral demand. Understanding unemployment patterns across different demographic and educational groups provides valuable insight into how opportunities are distributed and which populations face persistent challenges in the job market. \newline
  This research focuses on unemployment trends in the United States between 2020 and 2024, with particular attention to how education and gender influence employment outcomes. These factors are not only central to national labor policies but also directly relevant to college students entering the workforce for the first time. As many students navigate internships, job applications, and career planning, examining such patterns can offer a perspective on how education and demographic characteristics shape job prospects. Specifically, this paper investigates the following question: \newline
  
  **(1) Did college graduates have a lower average unemployment rate than non-college graduates betweeen 2020 and 2024?**

## Data Summary

  The data for unemployment rates of both college graduates and high school graduates come from the U.S. Bureau of Labor Statistics, retrieved from an article from FRED (Federal Reserve Bank of St. Louis). The U.S. Bureau of Labor Statistics collects data through a mix of methods, including surveys sent to households and businesses, electronic data interchange for large firms, and computer-assisted telephone interviews, to gather a variety of data such as employment, wages, and consumer expenditures. The data, also known as the Current Population Survey (CPS), is collected and processed by BLS employees and contractors with strict policies to ensure the confidentiality of respondent information. Further, the data collected for the CPS is a monthly sample survey of around 60,000 households from across the United States; the Census Bureau selected eligible households based on certain steps. They first group all of the counties and independent cities in the U.S. into 2,000 geographic areas, and design and select a sample of around 800 areas (units) to represent each state (and D.C.); these areas are intended to reflect a various range, including urban and rural areas, as well as various industrial and farming regions. The Census Bureau does note that their sample excludes individuals living in institutions and those on active duty in the Armed Forces. However, overall, any person over the age of sixteen is eligible to be selected for the survey (U.S. Bureau of Labor Statistics, n.d.-a). \newline
  According to the Labor Force Statistics from the CPS, much of the data collected can often undergo seasonal adjustments to account for fluctuations over the course of year as a result of seasonal events. However, both of the datasets used for this paper have not undergone any form of seasonal adjustments. The data does experience modifications for population control. Specifically, the Census Bureau adjusts the CPS population controls to incorporate the latest information about population change, and these adjustments are incorporated into the CPS estimates for data in January (U.S. Bureau of Labor Statistics, n.d.-b). \newline
  One of the main potential issues the data could pose is regarding its status as sample data. While it may be designed to be nationally representative, a sample is not a complete count of the U.S. population, meaning it is subject to random variation and sampling error. The impact of this is that reported unemployment rates can differ from the true population value, particularly for small subgroups like education level, which may affect comparison between the two groups. Additionally, there is an issue with how the CPS defines unemployment. According to the BLS, a person who is unemployed is characterized as someone who is jobless, looking for a job, and available for work (U.S. Bureau of Labor Statistics, 2015). This definition can exclude a large number of individuals, including discouraged workers who are unemployed but are not actively seeking employment. As a result, the collected unemployment rates may underrepresent the true values for the two groups.\newline
  Despite this, the collected data remains appropriate for answering our outlined research question, largely because the CPS is a major primary source for U.S. labor force statistics. It collects detailed data that distinguishes between the levels of educational attainment (those with a bachelor’s degree or higher vs. those with a high school degree or lower), allowing for a direct comparison of unemployment rates between the two groups. Additionally, because this data is collected monthly, it can successfully break down the single year that the research question focuses on (2024) with by-month data. 

## Data Dictionary

```{r data_dictionary_q1, echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)

# create the data frame
dd <- data.frame(
  `Variable Name` = c("Month", "URC", "URNC"),
  Description = c(
    "Calendar month (Jan–Dec, 2024); each row represents one observation.",
    "Monthly unemployment rate for individuals aged 25+ with a bachelor's degree or higher (unemployed ÷ labor force in this group).",
    "Monthly unemployment rate for individuals aged 25+ with a high school diploma only (unemployed ÷ labor force in this group)."
  ),
  `Unit / Type` = c(
    "Categorical (text)",
    "Percent (%) – Continuous",
    "Percent (%) – Continuous"
  ),
  check.names = FALSE
)

# print a simple bordered table
knitr::kable(dd, booktabs = TRUE, caption = "Data Dictionary – Q1")
```

## Data Exploration

### Graphical Summaries: Line Plot and Bar graph
```{r}


# Line Plot of Unemployment

ggplot(unemployment.data, aes(x=Month)) +
  geom_line(aes(y=UR_College, color="College", group=1)) +
  geom_line(aes(y=UR_NonCollege, color = "Non-College", group=1)) + 
  geom_point(aes(y=UR_College, color="College")) +
  geom_point(aes(y=UR_NonCollege, color = "Non-College")) + 
  labs(
    title = "Unemployment Rates in 2024: College vs. Non-College Graduates", 
    x = "Month",
    y = "Unemployment Rate (%)", 
    color = "Group") + 
  theme_minimal()

# Bar Graph
ggplot(unemployment.data, aes(x=Month)) + 
  geom_col(aes(y=UR_College, fill = "College"), position = "dodge") + 
  geom_col(aes(y=UR_NonCollege, fill = "Non-College"), position = "dodge") + 
  geom_text(aes(y=UR_College, label = UR_College), vjust = -0.5, color="black") + 
  geom_text(aes(y=UR_NonCollege, label = UR_NonCollege), vjust = -0.5, color="black") + 
  labs(
    title = "Unemployment Rates in 2024 with Numeric Values", 
    x = "Month",
    y = "Unemployment Rate (%)", 
    fill = "Group") + 
  theme_minimal()

```
### Numerical Summary: Five-Number Summary

```{r, echo=FALSE}
Summary.data <- summary(unemployment.data[,c("UR_College", "UR_NonCollege")])
Summary.data
```

### Average Unemployment by Education Level

```{r, echo=FALSE}
avg_college <- mean(unemployment.data$UR_College)
avg_noncollege <- mean(unemployment.data$UR_NonCollege)

averages <- c(College = avg_college, NonCollege = avg_noncollege)
averages
```

## Summary of the findings:

In the observed 2024 unemployment data, college graduates experienced higher and more volatile unemployment rates compared to non college workers. The average unemployment rate among college graduates was approximately 6.4% with values ranging from 4.1% to 9.0%. On the other hand, non college workers had a steadier trend averaging around 4.2% with a narrower range between 3.8% to 4.8%. The graph reinforce this pattern which shows the month to month unemployment for college graduates rose sharply at several points (in February and June) while unemployment among non college workers remained relatively stable throughout the year. \newline       
     Contrary to the traditional expectations that higher education ensures better job security, this dataset implies that labor market conditions in 2024 were less stable for college educated workers which shows the ongoing structural adjustments in white collar industries following post pandemic realignments. Based on these observations, Question 1: the relationship between educational attainment and unemployment stability appears most interesting for deeper statistical analysis as it highlights shifting dynamics in how education influences employment outcomes. 
     
\newpage

## Data Appendix

```{r}
head(unemployment.data, 12)
```

\newpage

## References 

- Federal Reserve Bank of St. Louis. (2000, January 1). Unemployment Rate - College Graduates - Bachelor’s Degree, 20 to 24 years. FRED, Federal Reserve Bank of St. Louis. https://fred.stlouisfed.org/series/CGBD2024
- Federal Reserve Bank of St. Louis. (2021, March 5). Unemployment Rate - High School Graduates, No College, 25 Yrs. & Over. Fred.stlouisfed.org. https://fred.stlouisfed.org/series/LNU04027660
- U.S. Bureau of Labor Statistics. (n.d.-a). Current Population Survey : U.S. Bureau of Labor Statistics. Www.bls.gov. https://www.bls.gov/cex/cecomparison/cps_profile.htm
- U.S. Bureau of Labor Statistics. (n.d.-b). Technical Documentation (CPS) : U.S. Bureau of Labor Statistics. Www.bls.gov. https://www.bls.gov/cps/documentation.htm
- U.S. Bureau of Labor Statistics. (2015, October 8). How the Government Measures Unemployment. Bls.gov. https://www.bls.gov/cps/cps_htgm.htm




